A finite-state morphological transducer for Kyrgyz

### 2011-10-19 dev/paper/paper.tex
things should make better now
some additions

\documentclass[10pt, a4paper]{article}

\usepackage{lrec2006}

@@

%FIXME: Should we say something about how I sometimes consulted Tolgonay, or about where my knowledge of Kyrgyz comes from in the first place?
The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \cite{hebertpoppe1963}, \cite{usonalievomuraliev2003}, \cite{qudaybergenov1980}, \cite{somfaikara2003}, and \cite{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \cite{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

@@

%FIXME: do we want subsections?
\subsection{Irregular negatives of finite verb forms}

@@

\begin{table}[htbp]
	\caption{examples of different affirmative / negative alternations in finite verb forms}\label{irregnegs}
	\centering
	\begin{tabular}{lll}
		\toprule
		tense/aspect & ending $+$ person series & ending $+$ person series \\
		\midrule
		recent eyewitness past & -/DI/ $+$ P4\footnotemark{} & -/GAн жок/ $+$ P3\\
		non-recent past & -/GAн/ $+$ P3 & -/GAн эмес/ $+$ P3\\
		non-recent evidential past & -/GAн экен/ $+$ P3 & -/GAн эмес экен/ $+$ P3\\
		past habitual & -/чU/ $+$ P3 & -/чU эмес/ $+$ P3\\
		recent evidential past & -/(I)птIр/ $+$ P3 & -/BAптIр/ $+$ P3\\
		habitual/future & -/E/ $+$ P6 & -/BAй/ $+$ P6\\
		%FIXME: /й/ here is technically /E/, but it's always realised as [й] after a vowel so will always be [й] here; how should we talk about it here?  Is this actually treated as regular in kymorph?
		\bottomrule
	\end{tabular}
\end{table}

@@

%FIXME: do we want subsections?
\subsection{Irregular [noun + possessive + case] forms}

@@

\begin{table}[htbp]
	\caption{combinations of possessive suffixes with case suffixes}\label{irregposcase}
	\begin{tabular}{llllll}
		\toprule
		case & morphology & 1st person singular & 2nd person sing. & 3rd person & 1st person plural \\
		\midrule
		nom & — & -(I)м & -(I)ң & -(S)I & -(I)бIз \\
		acc & -NI & -(I)мдI & -(I)ңдI & -\textbf{(S)Iн} & -(I)бIздI \\
		gen & -NIн & -(I)мдIн & -(I)ңдIн & -(S)IнIн & -(I)бIздIн \\
		loc & -DA & -(I)мдA & -(I)ңдA & -\textbf{(S)IндA} & -(I)бIздA \\
		abl & -DAн & -(I)мдAн, -\textbf{(I)мAн} & -(I)ндAн, -\textbf{(I)ңAн} & -\textbf{(S)IнAн} & -(I)бIздAн \\
		dat & -GA & -\textbf{(I)мA} & -\textbf{(I)ңA} & -\textbf{(S)IнA} & -(I)бIзгA \\
		\bottomrule
	\end{tabular}
\end{table}

@@

%FIXME: do we want subsections?
\subsection{Vowel harmony}

@@

%FIXME: do we want subsections?
\subsection{Voicing assimilation}\label{devoicing}
In Kyrgyz, there are two basic processes affecting the realisation of consonants when they are adjacent to other consonants: voicing assimilation and desonorisation.

@@

%FIXME: do we want subsections?
\subsection{Desonorisation}
Desonorisation happens to /N/ when it follows any consonant, and to /L/ when it follows a consonant of equal or lower sonority (i.e., /л/, nasals (/м/, /н/, /ң/), and obstruents, but not /й/, /р/, or vowels).  The resulting surface consonant is [д] for both /N/ and /L/, unless it follows a voiceless consonant, in which case it surfaces as [т].  An example includes the accusative and plural suffixes:

@@

%FIXME: do we want subsections?
\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.  The implementation of this in twol is very straightforward.
%FIXME: what else needs to be said here?

%FIXME: do we want subsections?
\subsection{Nouns ending in /рн/, etc.}

@@

\begin{table}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,731\\
		Adjective & 924\\
		Verb & 896 \\
		Proper noun & 745\\
		Adverb & 213\\
		Numeral & 50\\
		Postposition & 42\\
		Conjunction & 36\\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total:     & 7,333\\
		\bottomrule
	\end{tabular}
	\caption{FIXME}\label{table:stems}
	%FIXME: are we putting captions above or below tables?  I like them better below, but I got the impression that there's some standard to have them above.  Maybe it was just UW being table Nazis for my MA thesis.
\end{table}

@@

\begin{table}
	\centering
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known     & Naïve coverage    & Mean ambiguity\\
		\midrule
		Kyrgyz Wikipedia & 329,524   & 269,573   & 81.84 $\pm$ 3.23  & 4.54\\
		RFERL Kyrgyzstan & 4,107,555 & 3,586,000 & 87.37 $\pm$ 1.23  & 4.46\\
		\bottomrule
	\end{tabular}
	\caption{FIXME}
	\label{table:cov}
\end{table}

@@

%FIXME: * Some remaining transducer issues

\bibliographystyle{lrec2006}
\bibliography{paper}

>>>>

\documentclass[10pt,a4paper,twocolumn]{article}

%old lrec stuff doesn't work well =(
%\usepackage{lrec2006}

@@

%FIXED: Should we say something about how I sometimes consulted Tolgonay, or about where my knowledge of Kyrgyz comes from in the first place?
%FIXME: Tolgonay's name to acknowledgements (where?) for helping with Kyrgyz
The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \cite{hebertpoppe1963}, \cite{usonalievomuraliev2003}, \cite{qudaybergenov1980}, \cite{somfaikara2003}, and \cite{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \cite{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

@@

\subsection{Irregular negatives of finite verb forms}

@@

\begin{table*}[htbp]
	\caption{examples of different affirmative / negative alternations in finite verb forms}\label{irregnegs}
	\centering
	\begin{tabular}{lll}
		\toprule
		tense/aspect & ending $+$ person series & ending $+$ person series \\
		\midrule
		recent eyewitness past & -/DI/ $+$ P4\footnotemark{} & -/GAн жок/ $+$ P3\\
		non-recent past & -/GAн/ $+$ P3 & -/GAн эмес/ $+$ P3\\
		non-recent evidential past & -/GAн экен/ $+$ P3 & -/GAн эмес экен/ $+$ P3\\
		past habitual & -/чU/ $+$ P3 & -/чU эмес/ $+$ P3\\
		recent evidential past & -/(I)птIр/ $+$ P3 & -/BAптIр/ $+$ P3\\
		habitual/future & -/E/ $+$ P6 & -/BAй/ $+$ P6\\
		%FIXME: /й/ here is technically /E/, but it's always realised as [й] after a vowel so will always be [й] here; how should we talk about it here?  Is this actually treated as regular in kymorph?
		\bottomrule
	\end{tabular}
\end{table*}

@@

\subsection{Irregular [noun + possessive + case] forms}

@@

\begin{table*}[htbp]
	\caption{combinations of possessive suffixes with case suffixes}\label{irregposcase}
	\centering
	\begin{tabular}{llllll}
		\toprule
		case & morphology & 1st person singular & 2nd person sing. & 3rd person & 1st person plural \\
		\midrule
		nom & — & -(I)м & -(I)ң & -(S)I & -(I)бIз \\
		acc & -NI & -(I)мдI & -(I)ңдI & -\textbf{(S)Iн} & -(I)бIздI \\
		gen & -NIн & -(I)мдIн & -(I)ңдIн & -(S)IнIн & -(I)бIздIн \\
		loc & -DA & -(I)мдA & -(I)ңдA & -\textbf{(S)IндA} & -(I)бIздA \\
		abl & -DAн & -(I)мдAн, -\textbf{(I)мAн} & -(I)ндAн, -\textbf{(I)ңAн} & -\textbf{(S)IнAн} & -(I)бIздAн \\
		dat & -GA & -\textbf{(I)мA} & -\textbf{(I)ңA} & -\textbf{(S)IнA} & -(I)бIзгA \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\subsection{Vowel harmony}

@@

\subsection{Voicing assimilation}\label{devoicing}
In Kyrgyz, there are two basic processes affecting the realisation of consonants when they are adjacent to other consonants: voicing assimilation and desonorisation.

@@

\subsection{Desonorisation}
Desonorisation happens to /N/ when it follows any consonant, and to /L/ when it follows a consonant of equal or lower sonority (i.e., /л/, nasals (/м/, /н/, /ң/), and obstruents, but not /й/, /р/, or vowels).  The resulting surface consonant is [д] for both /N/ and /L/, unless it follows a voiceless consonant, in which case it surfaces as [т].  An example includes the accusative and plural suffixes:

@@

\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.  The implementation of this in twol is very straightforward.
%FIXME: what else needs to be said here?

\subsection{Nouns ending in /рн/, etc.}

@@

\begin{table*}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,731\\
		Adjective & 924\\
		Verb & 896 \\
		Proper noun & 745\\
		Adverb & 213\\
		Numeral & 50\\
		Postposition & 42\\
		Conjunction & 36\\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total:     & 7,333\\
		\bottomrule
	\end{tabular}
	\caption{FIXME}\label{table:stems}
	%FIXME: are we putting captions above or below tables?  I like them better below, but I got the impression that there's some standard to have them above.  Maybe it was just UW being table Nazis for my MA thesis.
\end{table*}

@@

\begin{table*}[htbp]
	\centering
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known     & Naïve coverage    & Mean ambiguity\\
		\midrule
		Kyrgyz Wikipedia & 329,524   & 269,573   & 81.84 $\pm$ 3.23  & 4.54\\
		RFERL Kyrgyzstan & 4,107,555 & 3,586,000 & 87.37 $\pm$ 1.23  & 4.46\\
		\bottomrule
	\end{tabular}
	\caption{FIXME}
	\label{table:cov}
\end{table*}

@@

%FIXME: * Some remaining transducer issues
Also, despite the good coverage, there are still a number of grammatical forms that have not been implemented into the transducer.  One of these is the irregular combination of 3rd person non-past finite verb plus the yes/no question particle.  For example, the third person non-past of /бар/ \eng{go} is [барат], but when the question particle -/BI/ is added, the resulting form is [барабы] instead of *[баратпы] (the transducer's current output).

\bibliographystyle{styler}
\bibliography{paper}
